\begin{exercises}

\ex{10.1.1}
Draw and identify each of the curves
defined by the following parametrizations.
\begin{exenum}
\x
$P(t) = (t, t^2)$, \quad $-\infty < t < \infty$.
\x
$P(t) = (t-1, t^2)$, \quad $-\infty < t < \infty$.
\x
$P(t) = (t^2-1, t+1)$, \quad $-\infty < t < \infty$.
\x
$P(t) = (2t^{\frac13}, 3t^{\frac13})$
\quad $-\infty < t < \infty$.
\x
$P(t) = (t-1, t^3)$, \quad $-\infty < t < \infty$.
\x
$P(t) = (3\cos t, 3\sin t)$, \quad $0 \leq t \leq \pi$.
\x
$P(s) = (\sin s, 2)$, \quad $-\infty < s < \infty$.
\x
$Q(r) = (2\sin r, 3\cos r)$, \quad $-\infty < r < \infty$.
\end{exenum}

\ex{10.1.2}
Draw and identify each of the following
parametrized curves.
\begin{exenum}
\x
$\dilemma{x=t-1,}
{y=2t+3, & -\infty < t < \infty.}$
\x
$\dilemma{x=t^2,}
{y=t-3, & -\infty < t < \infty.}$
\x
$\dilemma{2 \cos t,}
{y=\cos t, & -\infty < t < \infty.}$
\x
$\dilemma{x=3\sec\theta,}
{y=2\tan\theta, & -\frac{\pi}2 < t < \frac{\pi}2.}$
\end{exenum}

\ex{10.1.3}
Each of the following parametrized curves
is a function $f$ of $x$.
[To put it another way, each is the graph
of an equation $y = f(x)$.]
Find $f(x)$.
\begin{exenum}
\x
$P(t) = (t-1, t^2+1)$, \quad $-\infty < t < \infty$.
\x
$\dilemma{x=t,}{y=e^{t^2}, & -\infty < t < \infty.}$
\x
$\dilemma{x=2\cos t}{y=3\sin t, & 0 \leq t \leq \pi.}$
\x
$P(t) = (e^t, t)$, \quad $-\infty < t < \infty$.
\end{exenum}

\ex{10.1.4}
For each of the following parametrization,
find an equation $F(x,y) = c$ whose
graph is the parametrized curve
\begin{exenum}
\x
$P(t) = (t^2,t)$, \quad $-\infty < t < \infty$.
\x
\xlab{10.1.4b}
$\dilemma{x=e^{3t},}{y=e^t, & -\infty < t < \infty.}$
\x
\xlab{10.1.4c}
$\dilemma{x=e^t+e^{-t},}
{y=e^t-e^{-t}, & -\infty < t < \infty.}$
\end{exenum}
[For \exref{10.1.4b} and \exref{10.1.4c}, you
will need in addition to the equation
$F(x,y) = c$, the inequality $x>0$.]

\ex{10.1.5}
For the ellipse in Example \exampref{10.1.2}
parametrized by the equations $x=4 \cos t$
and $y=3 \sin t$, interpret $t$ geometrically.
(\emph{Hint:} See Figure \figref{10.5}.)

\ex{10.1.6}
Sketch the curve defined by the parametrization
\[
\dilemma{x=a\theta - b\sin\theta,}
{y=a-b\cos\theta, & -\infty < \theta < \infty.}
\]
This curve is traced by a point, fixed on a
radius of a wheel of radius $a$ at a
distance $b$ from the center, as the wheel
rolls along a straight line.
There are two cases.
\begin{exenum}
\x
The \dt{curtate cycloid}, where $a>b$.
(Think of a point on the spoke of a wheel.)
\x
The \dt{prolate cycloid}, where $a<b$.
(Think of a point on the flange of a railway wheel.)
\end{exenum}

\ex{10.1.7}
For each of the following parametrized curves,
assume that $y$ is defined as a differentiable
function of $x$ in a neighborhood of the points
indicated, and find $\dydx$ at the point.
\begin{exenum}
\x
$P(t) = (2t+1, t^2)$, when $t=2$.
\x
$\dilemma{x=5 \cos s,}
{y=3 \sin s, \mbox{when} \: s = \frac{\pi}4.}$
\x
$\dilemma{x=e^t,}
{y=t, \mbox{when} \: t=0.}$
\x
$\dilemma{x=e^t,}
{y=t, \mbox{when} \: t = \ln 5.}$
\end{exenum}

\ex{10.1.8}
Find the slope of each of the following
parametrized curves at the point indicated.
\begin{exenum}
\x
$P(t) = (t-1, t^3-3t^2+3t-1)$, at $P(1)$.
\x
$\dilemma{x=3 \cos t,}
{y=3 \sin t, \mbox{when} \: t = \frac{\pi}4.}$
\x
$\dilemma{x = t^3 - t + 1,}
{y = t^2 + t +1, \mbox{at} \: (1, 3).}$
\x
$Q(t) = (t^2 - t + 1, e^t + 1)$, at $Q(0)$.
\end{exenum}

\ex{10.1.9}
\begin{exenum}
\x
Assuming the necessary differentiability
conditions on the parametrization
$P(t) = (x(t), y(t))$, derive a formula for
$\deriv2y$
\x
For the curve parametrized by the equations
$x = t^2 + t + 1$ and $y = t^3 + 3t$,
find $\deriv2y$ when $t=1$.
\end{exenum}

\end{exercises}
