\begin{exercises}

\ex{11.4.1}
Find the general real-valued solution of each
of the following
differential equations.
\begin{exenum}
\x
$\deriv2y - 6\dydx - 7y = 0$
\x
$\deriv2y + 6\dydx +9y = 0$
\x
$(D^2+6D+5)y = 0$
\x
$(D^2-2D+10)y = 0$
\x
$(4D^2+4D-3)y = 0$
\x
$\deriv2y + 6\dydx = 0$
\x
$(D^2+2D+6)y=0$
\x
$4\deriv2y + 4\dydx + y = 0$
\x
$\nxder2yt + 4 \nxder{}yt + 5y = 0$
\x
$\nxder2xt + \nxder{}xt + x = 0$.
\end{exenum}

\ex{11.4.2}
Show by substitution that the function
defined by $y=(c_1x + c_2)e^{rx}$
is a solution of the differential equation
$(D-r)^2y=0$.

\ex{11.4.3}
Let $r$ be a real number, and
$c_1$ and $c_2$ complex numbers.
Prove that, if $(c_1x+c_2)e^{rx}$ is a
real-valued function, then $c_1$ and $c_2$
must both be real.

\ex{11.4.4}
Let $r_1 =\alpha+i\beta$ and $r_2=\alpha-i\beta$,
where $\alpha$ and $\beta$ are real numbers
and $\beta \ne 0$.
Prove that, for any two complex numbers $c_1$
and $c_2$, if the function
\[
c_1e^{r_{1^x}} + c_2e^{r_{2^x}}
\]
is real-valued, then $c_1$ and $c_2$ are complex
conjugates of each other.

\ex{11.4.5}
For each of the following differential equations,
find the general real-valued solution by first
finding an equivalent set of first-order linear differential
equations and then solving these successively
to find $y$.
\begin{exenum}
\x
$(D+1)(D-2)(D-3)y=0$.
\x
$(D-2)(D^2-6D+9)y=0$.
\x
$(D-a)(D-b)(D-c)y=0$,
where $a$, $b$, and $c$ are distinct real numbers.
\end{exenum}

\ex{11.4.6}
Find the general real-valued solution of the
differential equation
\[
(D^2+4)(D-3)y=0
\]
by solving an equivalent pair of equations.
Use the fact that we have already derived
the general real-valued solution of the
second-order, homogeneous, linear
differential equation with constant coefficients.

\ex{11.4.7}
Using Theorem \thref{11.4.4}, which gives
the general real-valued solution of the $n$th-order
differential equation $p(D)y=0$,
solve each of the following.
\begin{exenum}
\x
$(D-2)(D+1)^2y=0$
\x
$\deriv3y - 7\dydx + 6y = 0$
\x
$(D-3)^2(D+1)(D-5)y=0$
\x
$D(D^2+3D-4)y=0$
\x
$(D+2)^3(D-1)y=0$
\x
$(D+3)^2(D^2+3)y=0$
\x
$\deriv3y + \deriv2y - 2\dydx = 0$
\x
$(D^2+2D+2)^2y=0$
\x
$(D+1)(D^2+2D+2)^2y=0$
\x
$D^2(D^2+2D+2)^2y=0$
\x
$\deriv4y - 81y = 0$
\x
$\deriv3y+\deriv2y+\dydx+y=0$.
\end{exenum}

\end{exercises}
