\begin{exercises}

\ex{1.6.1}
Let $f(x) = 3x^2 + 4$.
Using the definition of the derivative, compute
\begin{exenum}
\sx
$f^\prime(1)$
\sx
$f^\prime(a)$, for an arbitrary real number $a$.
\end{exenum}

\ex{1.6.2}
Write an equation of the line tangent to the graph of the
function $f^\prime$ in Problem \exref{1.6.1} at the point
\begin{exenum}
\sx
$(1,7)$
\sx
$(a, f(a))$.
\end{exenum}

\ex{1.6.3}
If $F(x) = \frac3{2x+1}$, compute $F^\prime(3)$ using the
definition of the derivative.

\ex{1.6.4}
Using the definition of the derivative, compute $f^\prime(a)$ for each
of the following functions.
\begin{exenum}
\sx
$f(x) = x^3$
\sx
$f(x) = x^2 + 3x +5$
\sx
$f(x) = 7$
\sx
$f(x) = \sqrt{x}, a > 0$
\sx
$f(x) = x + \frac1{x^2}$, $x \ne 0$
\sx
$f(x) = x^3 + 3x^2 + 3x + 1$
\sx
$f(x) = \sqrt{x^2 + 1}$
\sx
$f(x) = \frac1{\sqrt{x^2 + 1}}$
\sx
$f(x) = x^\frac13$.
\end{exenum}

\ex{1.6.5}
Using the results of Problem \exref{1.6.4}, find an equation of the line
tangent to the graph of $f$ at the point $(a, f(a))$, where
\begin{exenum}
\sx
$f(x) = x^3$ and $a = 0$.
\sx
$f(x) = x^2 + 3x + 5$ and $a=1$.
\sx
$f(x) = 7$ and $a$ is arbitrary.
\sx
$f(x) = x + \frac1{x^2}$ and $a$ is not zero.
\end{exenum}

\ex{1.6.6}
\begin{exenum}
\sx
If $F(x) = x^2$, use the definition of the derivative to find
$F^\prime(x)$.
\sx
Plot the graphs of $F$ and $F^\prime$ on the same $xy$-plane.
\end{exenum}

\ex{1.6.7}
\begin{exenum}
\sx
Show that the function $|x|$ is not differentiable at $0$
and interpret this fact geometrically.
\sx
Compute the derivative at $-1$ and at $1$ of the function $|x|$.
\end{exenum}

\ex{1.6.8}
Show that the function $\sqrt x$ is not differentiable at $0$.
Draw the graph and interpret the nondifferentiability geometrically.

\ex{1.6.9}
Using the results of Problem \exref{1.6.4}, find
\begin{exenum}
\sx
$\ddxof f (-1)$ if $f(x) = x^2 + 3x + 5$.
\sx
$\ddxof f (3)$ if $f(x) = x^3$.
\sx
$\ddxof f (b)$ if $f(x) = x^3 + 3x^2 + 3x + 1$.
\sx
$\ddxof{\sqrt{x^2 + 1}}$.
\sx
$\ddxof{(x^2 + 3x + 5)} (a)$.
\sx
$\ddx \left( x + \frac1{x^2} \right) $.
\end{exenum}

\ex{1.6.10}
\begin{exenum}
\sx
If $y = 2x + 1$, find $\dydx (a)$.
\sx
If $s = 16t^2$, find $\ddxof s (2)$.
\sx
If $s = 16t^2$, find $\ddxof s$.
\end{exenum}

\ex{1.6.11}
Using the definition of the derivative, prove that if
$y = ax^2 + bx + c$, then
$\dydx = 2ax + b$.

\ex{1.6.12}
Give an example of a continuous function that fails to have a derivative
at some point.

\end{exercises}
