\begin{exercises}

\ex{2.1.1}
Sketch the graph of each of the following functions,
carefully labeling all extreme points and all points of
inflection.  Classify each extreme point as to type.
\begin{exenum}
\sx
$x^2 - 5x + 6$
\sx
$3 - 2x - x^2$
\sx
$2x^2 - 3x - 1$
\sx
$5-2x^2$
\sx
$x^3 - 3x$
\sx
$(x+1)(x^3 - x^2 - 5x + 13)$
\sx
$\frac{x+2}x$
\sx
$\frac{16}x + x^2$
\sx
$x^3 - x$
\sx
$x^4 - x$
\sx
$(a^\frac23 - x^\frac23)^\frac32$
\sx
$|x|$
\sx
$|x-7|$
\sx
$3 + \frac6{x-2}$
\sx
$x^4 - 8x^2 + 3$
\sx
$\frac{x^2 - 5x + 4}{8x}$
\sx
$(x-1)(x-2)(x-3)$
\sx
$1+6x - \frac12x^3$.
\end{exenum}

\ex{2.1.2}
Show that a polynomial function of $x$ which consists
only of even powers of $x$ is an even function.

\ex{2.1.3}
Show that a polynomial function of $x$ which consists
only of odd powers of $x$ is an odd function.

\ex{2.1.4}
\begin{exenum}
\sx
Show that the graph of the function
$ax^2 + bx + c, a \ne 0$, always has an absolute extreme point.
\sx
Which of the constants $a$, $b$, and $c$ determines
the type of extreme point of the graph?
\sx
\sxlab{2.1.4c}
What is the extreme value of
$ax^2 + bx + c$?
\sx
Write $ax^2 + bx + c$ as $a \left( x^2 + \frac ba x \right) + c$,
complete the square on $x^2 + \frac ba x$ without changing
the function, and find the result of \exref{2.1.4c} algebraically.
\end{exenum}

\ex{2.1.5}
Show that the graph of $x^3 - 12x$ has a local maximum point
but no absolute maximum point and that it also has a local
maximum point which is strictly local.

\ex{2.1.6}
Sketch the graph of $f(x)$, if $f(0)=3$
and $f^\prime (x) = -1$ for all real values of $x$.

\ex{2.1.7}
Sketch the graph of $f(x)$, if $f(-1)=2$
and $f^\prime (x) = \frac12$ for all real values of $x$.

\ex{2.1.8}
Sketch the graph of $f(x)$, if $f(0)=0$
and $f^\prime (x) = x$ for all real values of $x$.

\ex{2.1.9}
Construct a function which has a local maximum point,
with local maximum point defined as in this section, but would
not have a local maximum if the definition were changed to demand
$f(a) > f(x), x \ne a$, instead of $f(a) \geq f(x)$.

\ex{2.1.10}
Graph the function $x^\frac13$ and show that it has a point of
inflection where neither the first nor the second
derivative exists.

\end{exercises}
