\begin{exercises}

\ex{2.3.1}
A ball is thrown upward with an initial velocity of $80$ feet per second.
Its distance above the ground $t$ seconds later is given by
$s = 80t - 16t^2$.
\begin{exenum}
\sx
Show that it reaches its highest point when it has zero velocity.
\sx
Show that its acceleration is constant.
\sx
For how many seconds is it going up?
\sx
How high does it go?
\sx
Show that it strikes the ground with a speed of $80$
feet per second.
\end{exenum}

\ex{2.3.2}
A particle moves on the $x$-axis and its position $t$ seconds
after it starts is given by
$x(t) = 4t^3 - 42t^2 + 135t - 100$.
Describe its motion.

\ex{2.3.3}
A particle moves on the $y$-axis and its position $t$ seconds
after it starts is given by $y(t) = 144t - 288 - 16t^2$.
Describe its motion.

\ex{2.3.4}
A particle moves on the $x$-axis and its position $t$ seconds after
it starts is given by
$x(t) = t^3 - 3t^2 + 5$.
Describe its motion.

\ex{2.3.5}
A particle moves on the $y$-axis and its position $t$ seconds after
it starts is given by $y(t) = 3t^3 - 9t + 10$.
Describe its motion.

\ex{2.3.6}
A particle moves on the ellipse with equation
$4x^2 + 9y^2 = 36$.  When it is passing through the point
$(-3,0)$, what is the horizontal component of its velocity?

\ex{2.3.7}
A particle moves on the circle with equation $x^2 + y^2 = 16$.
Show that $\frac{v_y}{v_x}$ at $(a,b)$ is equal to the slope
of the tangent to the circle at $(a,b)$.

\ex{2.3.8}
Two ships leave the same dock at noon, one traveling due north
at $15$ knots (nautical miles per hour) and the other due east
at $20$ knots.  At what rate it the distance between them increasing
at $2$ \textsc{p.m.}?

\ex{2.3.9}
Water is pouring into a conical funnel and, although it is also
running out of the bottom, the amount of water in the funnel
is increasing at the rate of $3$ cubic inches per minute.
If the conical part of the funnel is $5$ inches deep and the mouth
of the funnel is $6$ inches in diameter, how fast is the water
rising when it is $2$ inches deep?

\ex{2.3.10}
\begin{exenum}
\sx
Show that, at any instant, the ratio of the rate at which the area
of a circle is changing to the rate at which the radius is changing
is equal to the circumference of the circle.
\sx
Show that, at any instant, the ratio of the rate at which the volume
of a sphere is changing to the rate at which the radius is changing
is equal to the surface area of the sphere.
\end{exenum}

\ex{2.3.11}
A man $6$ feet tall walks away from a lamppost $15$ feet tall
at a rate of $4$ miles per hour.  How fast is his shadow lengthening
when he is $12$ feet from the pole?  How fast is the distance from
the foot of the lamppost to the tip of his shadow lengthening?

\ex{2.3.12}
At $3$ \textsc{p.m.} a ship which is sailing due south at $12$ knots
is $5$ miles west of a west-bound hip which is making $16$ knots.
\begin{exenum}
\sx
At what rate is the distance between the ships changing at $3$
\textsc{p.m.}?
\sx
At what time does the distance between the ships stop decreasing
and start increasing?
\sx
What is the shortest distance between the ships?
\end{exenum}

\ex{2.3.13}
A ladder $20$ feet long leans against a vertical wall.
The bottom of the ladder slides away from the wall at a constant
rate of $1$ foot per second.  At what rate is the top coming
down the wall when it it $12$ feet from the ground?

\ex{2.3.14}
A particle moves on the parabola with equation $y = x^2$.
The horizontal component of the velocity at each point is equal
to twice the abscissa of the point.  Show that the vertical component
of the velocity at each point is equal to
four times the ordinate of the point.

\ex{2.3.15}
Sand is being poured on the ground at a rate of $4$ cubic feet
per minute.  At each moment, it forms a conical point with the
height of the cone $\frac73$ of the radius of the base.
How fast is the height of the pile rising when $21\pi$ cubic feet
of sand is in the pile?

\end{exercises}
