\begin{exercises}

\ex{3.2.1}
Write an equation for each of the following.
\begin{exenum}
\sx
The parabola with focus at $(-2,0)$ and directrix $x=2$.
\sx
The parabola with focus at $(0,3)$ and directrix $y=-3$.
\sx
The parabola with focus at $(0,-1)$ and directrix $y=1$.
\sx
The parabola with focus at $(4,0)$ and vertex at the origin.
\sx
The parabola with focus at $(0,-2)$ and vertex at $(0,2)$.
\sx
The parabola with vertex at $(-5,0)$ and directrix $x+1=0$.
\end{exenum}

\ex{3.2.2}
Sketch the graph of each of the parabolas in Problem \exref{3.2.1}.

\ex{3.2.3}
Sketch on the same set of axes the graph of each of the
following equations.
Compare and contrast the graphs,
noting common features and differences.
\begin{exenum}
\sx
$y^2 = \frac12 x$
\sx
$y^2 = x$
\sx
$y^2 = 2x$
\sx
$y^2 = 3x$
\end{exenum}

\ex{3.2.4}
Sketch on the same set of axes the graph of each of the
following equations.
Compare and contrast the graphs,
noting common features and differences.
\begin{exenum}
\sx
$y^2 = 2x$
\sx
$y^2 = -2x$
\sx
$x^2 = 2y$
\sx
$x^2 = -2y$.
\end{exenum}

\ex{3.2.5}
Consider a point $(x_1,y_1)$ on the graph of
$y^2 = 4ax$.
\begin{exenum}
\sx
\sxlab{3.2.5a}
Find the slope of the tangent to the graph at $(x_1,y_1)$.
\sx
Write an equation of the tangent line in \exref{3.2.5a}.
\sx
Show that $yy_1 = 2a(x+x_1)$ is an equation of
the tangent line.
\end{exenum}

\ex{3.2.6}
Consider a point $(x_1,y_1)$ on the graph of $x^2 = 4ay$.
Show that $xx_1 = 2a(y+y_1)$ is an equation of the
line tangent to the graph at $(x_1,y_1)$.

\ex{3.2.7}
If $(x_1,y_1)$ lies on the graph of $y = ax^2 + bx + c$,
show that $\frac12(y + y_1) = axx_1 + \frac12b(x + x_1) + c$
is an equation of the line tangent to the graph at $(x_1,y_1)$.

\ex{3.2.8}
Write an equation of a line which passes through the point
$(8,7)$ and is a tangent to the graph of $y^2 = 6x$.

\ex{3.2.9}
\begin{exenum}
\sx
\sxlab{3.2.9a}
Find the point where the tangent to $y^2 = 4ax$
at the point $(x_1,y_1)$ cuts the $x$-axis.
Assume that $a \ne 0$.
\sx
Show that the segment of the tangent line between
$(x_1,y_1)$ and the point found in \exref{3.2.9a}
is bisected by the $y$-axis.
\end{exenum}

\ex{3.2.10}
Write an equation for each of the following:
\begin{exenum}
\sx
The parabola with vertex $(1,1)$ and directrix $x=-1$.
\sx
The parabola with vertex $(1,1)$ and directrix $y=0$.
\sx
The parabola with vertex $(4,3)$ and directrix $x=-2$.
\sx
The parabola with vertex $(-1,2)$ and directrix $y=4$.
\end{exenum}

\ex{3.2.11}
Find the focus of each of the parabolas in Problem \exref{3.2.10}.

\ex{3.2.12}
Do Problem \exref{3.2.10} with each occurence of the word
``vertex'' replaced by ``focus.''

\ex{3.2.13}
Find the focus, vertex, and directrix of the parabola which
is the graph of each of the following equations or functions.
Sketch the graph.
\begin{exenum}
\sx
$y = x^2$
\sx
$y = x^2 - 2x$
\sx
$f(x) = x^2 - 6x + 1$
\sx
$y^2 + 2y + 2x = 0$
\sx
$g(y) = -y^2 - 4y + 4$
\sx
$x^2 + x + y = 0$.
\end{exenum}

\end{exercises}
