\begin{exercises}

\ex{3.4.1}
Simplify $\sqrt{(x+c)^2 + y^2} - \sqrt{(x-c)^2 + y^2} = 2a$
and show that it results in the equation
$\frac{x^2}{z^2} - \frac{y^2}{c^2-a^2} = 1$.

\ex{3.4.2}
Describe and sketch the graph of each of the following equations.
Label the foci and vertices and sketch the asymptotes.
\begin{exenum}
\sx
$\frac{x^2}{9} - \frac{y^2}{16} = 1$
\sx
$\frac{x^2}{18} - \frac{y^2}{7} = 1$
\sx
$\frac{x^2}{25} - \frac{y^2}{144} = 1$
\sx
$\frac{x^2}{225} - \frac{y^2}{64} = 1$
\sx
$25y^2 - 9x^2 = 225$
\sx
$xy = 8$
\sx
$xy = -6$.
\end{exenum}

\ex{3.4.3}
It has been shown that the distance between a point on
the hyperbola $\frac{x^2}{a^2} - \frac{y^2}{c^2-a^2} = 1$
and the focus $(c,0)$ is $\left| \frac ca x - a \right|$.
Call this distance $d_1$.
\begin{exenum}
\sx
Show that the distance $d_2$ between a point on the
hyperbola and the focus $(-c,0)$ is
$\left| \frac ca x + a \right|$.
\sx
Show that $x \geq a$ for a point on the right branch of
the hyperbola and that for such a point
$d_1 = \frac ca x - a$ and $d_2 = \frac ca x + a$.
\sx
Show that $x \leq -a$ for a point on the left branch of
the hyperbola and that for such a point
$d_1 = -\frac ca x + a$ and $d_2 = -\frac ca x - a$.
\sx
Hence show that the graph of
$\frac{x^2}{a^2} - \frac{y^2}{c^2-a^2} = 1$
contains only those points which satisfy the locus
definition of hyperbola.
\end{exenum}

\ex{3.4.4}
Write an equation for the hyperbola with horizontal
and vertical axes satisfying the given conditions.
\begin{exenum}
\sx
Foci at $(-5,0)$ and $(5,0)$.
Transverse axis of length $6$.
\sx
Center at the origin.
Transverse axis horizontal and of length $4$,
conjugate axis vertical and of length $12$.
\sx
Center at the origin.
Transverse axis vertical and of length $6$.
Passing through $(1, 2\sqrt3)$.
\sx
Center at the origin.
Passing through $(1,5)$ and $(2,7)$.
\end{exenum}

\ex{3.4.5}
\begin{exenum}
\sx
\sxlab{3.4.5a}
Show that the graph of $(x-3)(y+2) = 10$ is an
equilateral hyperbola with center at
$(3,-2)$ and asymptotes $x=3$ and $y=-2$.
\sx
Sketch the graph described in \exref{3.4.5a}.
\end{exenum}

\ex{3.4.6}
Describe and sketch the graph of each of the following equations.
Label the foci and vertices and sketch the asymptotes.
\begin{exenum}
\sx
$\frac{(x-1)^2}{16} - \frac{(y+3)^2}{9} = 1$
\sx
$\frac{(y+3)^2}{9} - \frac{(x-1)^2}{16} = 1$
\sx
$(x-1)(y+3) = -12$
\sx
$144(x+4)^2 - 25(y+2)^2 = 3600$
\sx
$2xy - 4x - 4y - 17 = 0$
\sx
$x^2 - y^2 - 2x + 2y - 2 = 0$
\sx
$y^2 - 4x^2 + 4y - 16x - 28 = 0$.
\end{exenum}

\ex{3.4.7}
Hyperbolas with the same pair of foci are said to be
\dt{confocal}.  Show that the following equations
have confocal hyperbolas for their graphs and sketch
them all on the same set of axes.
\begin{exenum}
\sx
$x^2 - \frac{y^2}{24} = 1$
\sx
$\frac{x^2}{9} - \frac{y^2}{16} = 1$
\sx
$\frac{x^2}{16} - \frac{y^2}{9} = 1$
\sx
$\frac{x^2}{24} - y^2 = 1$.
\end{exenum}

\ex{3.4.8}
The line segment which passes through a focus,
is perpendicular to the transverse axis extended,
and has its endpoints on the hyperbola is called a
\dt{latus rectum}.
\begin{exenum}
\sx
Find the length of a latus rectum of the hyperbola
$9x^2 - 16y^2 = 144$.
\sx
Find the length of a latus rectum of the hyperbola
$b^2x^2 - a^2y^2 = a^2b^2$.
\sx
Show that both latera recta of a hyperbola are the same length.
\end{exenum}

\ex{3.4.9}
Write equations of the directrices and find the eccentricity
of each of the following hyperbolas.
\begin{exenum}
\sx
$9x^2 - 16y^2 = 144$
\sx
$144y^2 - 25x^2 = 3600$.
\end{exenum}

\ex{3.4.10}
Assume that $0 < a < c$.
\begin{exenum}
\sx
\sxlab{3.4.10a}
Find the distance between $(x,y)$ and $(-c,0)$.
\sx
\sxlab{3.4.10b}
Find the distance between $(x,y)$ and the line
$x = -\frac{a^2}c$.
\sx
Find the locus of points $(x,y)$ such that the ratio between
the distance in \exref{3.4.10a} and the distance in \exref{3.4.10b}
is a constant $\frac ca$.
\end{exenum}

\ex{3.4.11}
Consider points of the upper right branch of the hyperbola
$x^2-y^2=16$ and on its asymptote $x=y$.
Find the $y$-coordinates of the points on each for
$x = 10, 100, 1000, 10,000$, and show that the
vertical distance between corresponding points
is decreasing as $x$ increases.

\ex{3.4.12}
Consider a point $(x_1,y_1)$ on the graph of
$b^2x^2 - a^2y^2 = a^2b^2$.
\begin{exenum}
\sx
\sxlab{3.4.12a}
Find the slope of the tangent to the graph at
$(x_1,y_1)$.
\sx
Write an equation of the tangent line in \exref{3.4.12a}.
\sx
Show that
$b^2xx_1-a^2yy_1=a^2b^2$
is an equation of the tangent line.
\end{exenum}

\ex{3.4.13}
Consider a point $(x_1,y_1)$ on the hyperbola
$ax^2-by^2+cx+dy+e=0$ with $ab>0$.
\begin{exenum}
\sx
\sxlab{3.4.13a}
Find the slope of the tangent to the graph at
$(x_1,y_1)$.
\sx
Write an equation of the tangent line in \exref{3.4.13a}.
\sx
Show that
$axx_1-byy_1+\frac12c(x+x_1)+\frac12d(y+y_1)+e=0$
is an equation of the tangent line.
\end{exenum}

\ex{3.4.14}
Show that the product of the distances of a point on the
hyperbola $xy = -12$ to its asymptotes is a constant.

\ex{3.4.15}
If the difference of the distances of the point $(x,y)$
from two foci is zero, show that the locus of $(x,y)$
is the perpendicular bisector of the line segment
joining the foci.

\ex{3.4.16}
Describe and sketch the graph of each of the following
equations.  If the graph is a circle, give its center and focus.
If the graph is a parabola, give its focus, directrix, vertex,
and axis.  If the graph is an ellipse, give its center, foci,
directrices, eccentricity, and length of major and minor
axes.  If the graph is an hyperbola, give its center, foci,
directrices, eccentricity, asymptotes, vertices, and length
of transverse axis.
\begin{exenum}
\sx
$x^2 + y^2 + 6x + 4y = 12$
\sx
$x^2 + 4y^2 + 6x + 4y + 6 = 0$
\sx
$x^2 + 6x + 4y + 2 = 0$
\sx
$x^2 - 4y^2 + 6x + 4y + 4 = 0$
\sx
$4y^2 + 6x + 4y + 13 = 0$
\sx
$xy + 6x + 4y = 3$
\sx
$3x^2 + 3y^2 + 6x - 18y = 162$
\sx
$4y^2 + x^2 + 6x + 4y = 11$
\sx
$y^2 = 9x^2 + 2y + 8$
\sx
$y^2 = 2y -9x^2 + 8$
\end{exenum}

\end{exercises}
