\begin{exercises}

\ex{4.5.1}
Verify that $\frac1{r+1} x^{r+1}$ is an antiderivative of
$x^r$, if $r$ is any rational number except $-1$.

\ex{4.5.2}
Find an antiderivative of each of the following functions.
\begin{exenum}
\x
$f(x) = x^7$
\x
$f(x) = x^3 + \frac1{x^3}$
\x
$f(y) = 7y^\frac15$
\x
$f(t) = 5t^4 + 3t^2 + 1$
\x
$g(x) = (3x + 1)^2$
\x
$f(x) = \frac{2x}{(x^2 + 1)^2}$.
\end{exenum}

\ex{4.5.3}
Evaluate each of the following definite integrals by finding
an antiderivative and using Theorem \thref{4.5.3}.
\begin{exenum}
\x
$\int_0^1 3x^2 \; dx$
\x
$\int_0^1 (4x^3 + 3x^2 + 2x + 1) \; dx$
\x
$\int_1^3 (5x - 1) \; dx$
\x
$\int_1^3 (5t - 1) \; dt$
\x
$\int_1^2 \left( x^2 + \frac1{x^2} \right) \; dx$
\x
$\int_3^2 x^\frac13 \; dx$
\x
$\int_{-2}^0 y^\frac15 \; dy$
\x
$\int_1^2 \left( \frac2{x^3} + \frac1{x^2} + 2 \right) \; dx$
\x
$\int_{-1}^1 (y^2 - y + 1) \; dy$
\x
$\int_6^0 (x^3 - 9x^2 + 16x) \; dx$
\x
$\int_3^5 (2x - 1)^2 \; dx$
\x
$\int_3^x (6t^2 - 4t + 2) \; dx$
\x
$\int_0^t (x^2 + 3x - 1) \; dx$
\x
$\int_0^{x^2} s^3 \; ds$
\x
$\int_a^b dx$
\x
$\int_x^{3x} (4t - 1) \; dt$.
\end{exenum}

\ex{4.5.4}
Let $n$ be a positive integer.
\begin{exenum}
\x
Evaluate $\int_a^b x^n \; dx$.
\x
\xlab{4.5.4b}
Evaluate $\int_a^b \frac1{x^n} \; dx$ provided
(i) $n \ne 1$, and (ii) $a$ and $b$ are either
both positive or both negative.
\x
In \exref{4.5.4b}, what is the reason for proviso (i)?
For proviso (ii)?
\end{exenum}

\ex{4.5.5}
Let $F$ be an antiderivative of $f$ and $G$
an antiderivative of $g$.
\begin{exenum}
\x
Prove that $F+G$ is an antiderivative of $f+g$.
\x
For any constant $k$,
prove that $kF$ is an antiderivative of $kf$.
\end{exenum}

\ex{4.5.6}
What is the domain of the function $F$ defined by
$F(t) = \int_1^t \frac1x \; dx$.

\ex{4.5.7}
Let $F(t) = \int_0^t (6x^2 - 4x + 1) \; dx$.
\begin{exenum}
\x
\xlab{4.5.7a}
Using just the Fundamental Theorem and without evaluating
$F$, find $F^\prime(t)$, $F^\prime(-1)$, $F^\prime(2)$,
and $F^\prime(x)$.
\x
\xlab{4.5.7b}
Find $F(t)$ as a polynomial in $t$ by finding a polynomial
which is an antiderivative of $6x^2 - 4x + 1$.
\x
Differentiate the answer in \exref{4.5.7b},
and thereby check \exref{4.5.7a}.
\end{exenum}

\ex{4.5.8}
Let $G(x) = \int_1^x \left( t + \frac1{t^2} \right) \; dt$,
for $x > 0$.
\begin{exenum}
\x
\xlab{4.5.8a}
Using just the Fundamental Theorem,
find $G^\prime(x)$ and $G^\prime(2)$.
\x
\xlab{4.5.8b}
Evaluate $G(x)$ as a rational function of $x$ by finding
an antiderivative of $t + \frac1{t^2}$.
\x
Take the derivative of $G(x)$ as found in \exref{4.5.8b}
and thereby check \exref{4.5.8a}.
\end{exenum}

\ex{4.5.9}
\begin{exenum}
\x
\xlab{4.5.9a}
Evaluate $F(t) = \int_0^{t^2} (3x^2 + 1) \; dx$.
\x
Find $F^\prime(t)$ and $F^\prime(2)$
by taking the derivative of the answer to \exref{4.5.9a}.
\x
Find $F^\prime(t)$ directly using just the Fundamental
Theorem and the Chain Rule.
\end{exenum}

\ex{4.5.10}
If $f$ is continuous and $g$ is differentiable and if
$F(t) = \int_a^{g(t)} f(x) \; dx$, use the Fundamental
Theorem and the Chain Rule to show that
$F^\prime (t) = f(g(t))g^\prime(t)$.

\ex{4.5.11}
In each of the following integrals evaluate $F^\prime (t)$.
Do not attempt to first find an antiderivative.
\begin{exenum}
\x
$F(t) = \int_0^t \sqrt{1+x^3} \; dx$.
\x
$F(t)  = \int_t^1 \frac1{1+x^2} \; dx$.
\x
$F(t) = \int_0^{2t+1} \frac1{1+x^2} \; dx$.
\x
$F(t) = \int_t^{t^2} \frac1{x^2+x+1} \; dx.
\left(\emph{Hint:} \int_t^{t^2} f = \int_t^1 f + \int_1^{t^2} f. \right)$
\end{exenum}

\ex{4.5.12}
Is there anything wrong with the computation
\[
\int_{-1}^1 \frac1{x^2} \; dx = 
\left. -\frac1x \right|_{-1}^1 = (-1) - (1) = -2
?
\]
If so, what?

\ex{4.5.13}
In each of the following, find the area of the subset $P$
of the $xy$-plane bounded by the curve $y=f(x)$,
the $x$-axis, and the lines $x=a$ and $x=b$.
Sketch the curve and the subset $P$.
\begin{exenum}
\x
$f(x) = x^2+1$, $a = -1$, and $b = 2$.
\x
$f(x) = x^2+2x$, $a = 0$, and $b = 2$.
\x
$f(x) = \frac12x+1$, $a = 2$, and $b = 4$.
\x
$f(x) = x^3$, $a = 0$, and $b = 2$.
\x
$f(x) = x^3-2x^2+x$, $a = 0$, and $b = 2$.
\end{exenum}

\ex{4.5.14}
Find the derivative of the function $F$ defined by
$F(x) = \int_0^x \frac1{t^2+1} \; dt$.
Sketch the graph of $F$ using the techniques of curve
sketching discussed in Section \secref{2.1}.
Label and maximum, minimum, or critical points
and any points of inflection.
What is the domain of $F$?
$\left( \mbox{Do not attempt to find an explicit 
antiderivative of $\frac1{t^2+1}$.} \right)$

\end{exercises}
