\begin{exercises}

\ex{5.5.1}
Find the general solution of each of the
following differential equations.
\begin{exenum}
\x
$\dydx - 3y = 0$
\x
$\nxder{} yt = t(t^2 + 1)$
\x
$x\dydx + y = 0$
\x
$(x+1)y \dydx = (y^2+1)$
\x
$\dydx e^{t-y}$
\x
$xy \dydx = y^2 - 2$.
\end{exenum}

\ex{5.5.2}
For each of the following differential equations,
find the particular solution whose graph passes
through the point indicated.
\begin{exenum}
\x
$\dydx = -\frac yx$, passing through $(1,1)$.
\x
$2\dydx = 3y$, passing through $(0,5)$.
\x
$y\dydx = 18x^3$, passing through $(2,-9)$.
\x
$\dydx + \frac xy = 0$, passing through $(5,0)$.
\end{exenum}

\ex{5.5.3}
A curve defined by $y = f(x)$ has slope $m$
at every point $(x,y)$ given by
$m = 2y$.
If the curve passes through the point
$(0,-1)$, find $f(x)$.

\ex{5.5.4}
Find all solutions to the differential equation
$\dydx = -\frac xy$.
Sketch the graphs of the different solutions.

\ex{5.5.5}
Find all solutions to the differential equation
$\dydx = \frac xy$.
Sketch the graphs of the different solutions.

\ex{5.5.6}
Classify each of the following differential equations
as separable, linear, both, or neither.
\begin{exenum}
\x
$\ln y \dydx = \frac yx$
\x
$x^2 \dydx + y = e^x$
\x
$y \dydx + x + y = 0$
\x
$\nxder{}xt -7t = 0$
\x
$\nxder{}xt - 7x = 0$
\x
$\sqrt{y^2+1} \dydx + x^2y = 0$
\x
$\frac1x \nxder{}xt = 3$
\x
$\dydx + \frac xy = x^2$
\x
$\left( \dydx \right)^2 + 3y = 7x$.
\end{exenum}

\ex{5.5.7}
An alternative approach to solving the linear
differential equation
$\dydx + ky = 0$ is to write it as
$\dydx = -ky$.  The latter equation is similar to
$\dydx = y$, which has $e^x$ for a solution.
With this similarity in mind,
it is not hard to guess, and then verify,
that $y = e^{-kx}$ is a solution to the original equation.
The problem is now to show that \emph{every}
solution is a constant multiple of
$e^{-kx}$.  Prove this fact by assuming that $y = f(x)$
is an arbitrary solution of
$\dydx + ky = 0$ and then showing that the derivative
of the quotient $\frac{f(x)}{e^{-kx}}$ is zero.
(See Problem \exref{5.3.8}.)

\ex{5.5.8}
A radioactive substance has a half-life of
$10$ hours.  What fraction of an amount
of this substance decays in $15$ hours?

\ex{5.5.9}
If a certain population increases at a rate proportional
to the number in the population and it doubles in
$45$ years, in how many years is it multiplied by
a factor of $3$?

\ex{5.5.10}
Find the constant of proportionality relating a
radioactive substance and its rate of decay
if the substance has a half-life of $16$ hours.

\ex{5.5.11}
The number $y$ of bacteria in a culture grows
at a rate $\nxder{}yt$ proportional to the number
present.  If the number doubles in $3$ days
and there are $10^7$ bacteria present
at the beginning of the experiment,
how many are there after $24$ hours?

\ex{5.5.12}
A toy block lying on the floor is given a kick.
The resulting acceleration $v^\prime$
(which is negative) is equal to $-5v$.
If the kick gives it an initial velocity of $6$ feet per second,
how many seconds later is the velocity equal to
$2$ feet per second?

\ex{5.5.13}
A car sliding along a track slows down at a rate
proportional to its velocity.
If it has one-half its initial velocity after $30$
seconds, at what fraction of its initial
velocity is it traveling after $1$ minute?

\ex{5.5.14}
Let $a$ and $b$ be constants with $a \ne 0$.
Show that the differential equation
\begin{equation}
\label{eq5.5.12}
\dydx + ay = b
\end{equation}
reduces to $\nxder{}zx + az = 0$ if we let
$z = y - \frac ba$.
As a result, find the general solution of
\eqref{5.5.12}.

\ex{5.5.15}
Use the substitution described in Problem \exref{5.5.14}
to find the particular solution of the differential
equation $\dydx - 2y = 6$ which passes through
the point $(0,4)$.

\end{exercises}
