\begin{exercises}

\ex{8.1.1}
Find the average value $M_a^b(f)$
of $f$ on the interval $[a,b]$, where
\begin{exenum}
\x
$f(x) = x^2-2x+1$ and $[a,b] = [0,2]$.
\x
$f(x) = 2x^3$ and $[a,b] = [-1,1]$.
\x
$f(x) = \frac1x$ and $[a,b] = [1,2]$.
\x
$f(x) = \frac1x$ and $[a,b] = [1,n]$,
where $n$ is a positive integer.
\x
$f(x) = \sin x$ and $[a,b] = [0,\pi]$.
\x
$f(x) = \ln x$ and $[a,b] = [1,5]$.
\end{exenum}

\ex{8.1.2}
In each of the following find $M_a^b(f)$,
draw the graph of $f$, and superimpose
on the graph a rectangle with base $[a,b]$
and area equal to the area under the curve
$y=f(x)$ between $a$ and $b$.
\begin{exenum}
\x
$f(x) = x^2$, $a = -1$, and $b = 1$.
\x
$f(x) = x^3$, $a = 0$, and $b = 1$.
\x
$f(x) = 4-(x-1)^2$, $a = 0$, and $b = 3$.
\x
$f(x) = e^x$, $a=0$, and $b=2$.
\x
$f(x) = \cos x$, $a=0$, and $b=\frac{\pi}2$.
\end{exenum}

\ex{8.1.3}
Each of the propositions
\thref{8.1.2}, \thref{8.1.3}, \thref{8.1.4},
and \thref{8.1.5} corresponds to one of the
basic properties of the definite integral
as they are enumerated in Theorems
\thref{4.4.1} through \thref{4.4.5}.
In general, the proof of each is obtained by
checking the special case $a=b$
separately and then using the formula
\[
M_a^b(f) = \frac1{b-a} \int_a^b f(x)\;dx,
\quad \mbox{for $a<b$}
,
\]
together with the appropriate property
of the integral.
\begin{exenum}
\x
Prove \thref{8.1.2}
\x
Prove \thref{8.1.3}
\x
Prove \thref{8.1.5}.
\end{exenum}

\ex{8.1.4}
A stone dropped from a cliff $400$ feet high
falls to the bottom with a constant
acceleration equal to $32$ feet per second per second.
That is,
\[
a(t) = v^\prime(t) = s^{\prime\prime}(t) = 32
,
\]
where the direction of increasing $s$ is downward.
If the stone is dropped at time $t=0$,
find the time it takes to reach the bottom of the cliff,
and the mean velocity during the fall.

\ex{8.1.5}
A typist's speed over an interval from $t=0$
to $t=4$ hours increases as she warms up
and then decreases as she gets tired.
Measured in words per minute,
suppose that her speed is given by
$v(t) = 6[4^2-(t-1)^2]$.
Find her speed at the beginning, at the end,
her maximum speed, and her average speed
over the $4$-hour interval.
How many words did she type during the $4$
hours?

\ex{8.1.6}
A particle moves during the interval of time
from $t=1$ second to $t=3$ seconds
with a velocity given by
$v(t) = t^2+2t+1$ feet per second.
Find the total distance that the particle
has moved and also the average velocity.

\ex{8.1.7}
For each of the functions and intervals in
Problem \exref{8.1.2}, find a number $c$
such that $a<c<b$ and
$M_a^b(f) = f(c)$.

\ex{8.1.8}
An arbitrary linear function $f$ is defined by
$f(x) = Ax+B$ for some constants
$A$ and $B$.  Show that
\[
M_a^b(f) = \frac{f(a)+f(b)}2
.
\]

\ex{8.1.9}
Let $x(t)$ be the number of bacteria in a
culture at time $t$, and let $x_0=x(0)$.
The number grows at a rate proportional
to the number present, and doubles in a
time interval of length $T$.  Find an expression
for $x(t)$ in terms of $x_0$ and $T$,
and find the average number of bacteria
present over the time interval $[0,T]$.

\end{exercises}
