\begin{exercises}

\ex{8.7.1}
Classify each of the following integrals as proper,
improper and convergent, or improper and divergent.
Evaluate any which are convergent if an
indefinite integral can be found.
\begin{exenum}
\x
$\int_0^2 \frac3{x^{\frac23}} dx$
\x
$\int_1^2 \frac1{\sqrt{2-x}}dx$
\x
$\int_2^3 \frac1{(x-2)^2} dx$
\x
$\int_0^1 \frac1{x^2+x+1} dx$
\x
$\int_0^{\frac{\pi}2} \tan x \; dx$
\x
$\int_0^1 \frac{\sin x}x dx$.
\end{exenum}

\ex{8.7.2}
Classify each of the following integrals
and evaluate any which are not divergent.
\begin{exenum}
\x
$\int_2^{\infty} \frac1{x^3} dx$
\x
$\int_0^2 \frac1{x^3} dx$
\x
$\int_{-1}^{\infty} (x^2-x+1) \; dx$
\x
$\int_0^{\infty} xe^{-x^2} dx$
\x
$\int_1^{\infty} \sin x \; dx$
\x
$\int_{-\infty}^1 e^xdx$
\x
$\int_0^{\infty} \frac1{(x+2)^2} dx$
\x
$\int_0^1 x \ln x \; dx$.
\end{exenum}

\ex{8.7.3}
Show that the integral
$\int_0^1 \frac1{x^s} dx$ is
\begin{exenum}
\x
proper if $-\infty < s \leq 0$.
\x
improper and convergent if $0<s<1$.
\x
improper and divergent if $1 \leq s < \infty$.
\end{exenum}

\ex{8.7.4}
For what values of $s$ is the integral
$\int_1^{\infty} \frac{dx}{x^s}$ convergent,
and for what values is it divergent?
Give reasons for your answers.

\ex{8.7.5}
Classify each of the following integrals,
and evaluate any which are not divergent
if an indefinite integral can be found.
\begin{exenum}
\x
$\int_{-1}^1 \frac1{x^{\frac23}} dx$
\x
$\int_0^2 \frac1{(x-1)^{\frac13}} dx$
\x
$\int_0^1 \frac{\tan x}x dx$
\x
$\int_0^{\infty} \frac1{x^s} dx$
\x
$\int_{-infty}^0 \frac{dx}{(x-2)^2}$
\x
$\int_{-infty}^{infty} e^{-|x|} dx$
\x
$\int_2^{\infty} \frac1{\sqrt{x-2}} dx$
\x
$\int_0^2 \frac1{(x+1)(x-1)} dx$
\x
$\int_1^{\infty} \frac{\ln x}x dx$
\x
$\int_1^{\infty} \frac{\ln x}{x^2} dx$.
\end{exenum}

\ex{8.7.6}
Prove Theorem \thref{8.7.1}.

\ex{8.7.7}
Using the Comparison Test for Integrals if necessary,
classify each of the following integrals.
\begin{exenum}
\x
$\int_{-\infty}^0 e^{-x^2} dx$
\x
$\int_1^{\infty} \frac1{x^2} \sin x \; dx$
\x
$\int_0^{\infty} e^{-x} \sin x \; dx$
\x
$\int_3^{\infty} \frac1{\sqrt{(x-1)(x-2)}} dx$
\x
$\int_0^1 x \sin \frac1x dx$
\x
$\int_0^1 \frac1{\sqrt{(x-1)(x-2)}} dx$.
\end{exenum}

\ex{8.7.8}
If $F(t) = \int_{-\infty}^t e^{-x^2} dx$,
find $F^\prime(0)$ and $F^\prime(1)$.

\ex{8.7.9}
\begin{exenum}
\x
Show that the area of the region $P$ bounded by
the $x$-axis, the line $x=1$, and the curve
$y=\frac1x$ is infinite.
\x
Show that the volume of the solid of revolution
obtained by rotation the region $P$
about the $x$-axis is finite.
\end{exenum}

\ex{8.7.10}
Prove that if $f$ is bounded on $(a,b]$ and integrable
over $[t,b]$ for every $t$ in $(a,b]$, then $f$
is integrable over $[a,b]$ and
$\lim_{t\goesto a+} \int_t^b f = \int_a^b f$.
[\emph{Hint:} The argument is essentially the same
as that in the proof of Theorem \thref{8.6.1}.]

\end{exercises}
