\begin{exercises}

\ex{9.1.1}
Determine whether or not each of the following
sequences converges, and evaluate
the limit if it does.
\begin{exenum}
\x
$a_n = \frac{n^2-1}{n^2+1}$
\x
$b_k = \frac{k^2}{(k+1)^2}$
\x
$s_n = \frac{n+1}{n^2}$
\x
$s_n = (-1)^n \frac{n+1}{n^2}$
\x
$a_i = \left( 1 + \frac1i \right)^i$
\x
$s_i = i \sin \frac1i$.
\end{exenum}

\ex{9.1.2}
Evaluate each of the following limits.
\begin{exenum}
\x
$\lim_{n\goesto\infty} \frac{(n+1)(n+3)}{n^2+3}$
\x
$\lim_{i\goesto\infty} e^{\frac1i}$
\x
$\lim_{k\goesto\infty} \frac{k}{\sqrt{k+1}}$
\x
$\lim_{n\goesto\infty} \cos n$
\x
$\lim_{n\goesto\infty} \frac1n \cos n$
\x
$\lim_{k\goesto\infty} \left(\sqrt{k} - \sqrt{k+1}\right)$.
\end{exenum}

\ex{9.1.3}
Determine whether or not each of the following
sequences $\{ s_n \}$ converges, and, if it does,
evaluate the limit.
\begin{exenum}
\x
$\cond{s_n = (-1)^n, & n=1,2,\ldots.}$
\x
$s_n = \dilemma{1+\frac1n, & \mbox{for every integer
$n$ such that $1 \leq n \leq 10$,}}
{1, & \mbox{for every integer $n>10$.}}$
\x
$s_n  = \dilemma{1+\frac1n, & \mbox{if $n$ is
a positive even integer,}}
{1, & \mbox{if $n$ is a positive odd integer.}}$
\x
$s_n = \dilemma{1+\frac1n, & \mbox{for every
integer $n$ such that $1 \leq n \leq 10$,}}
{2, & \mbox{for every integer $n > 10$.}}$
\end{exenum}

\ex{9.1.4}
Let $s$ be the sequence defined by
$s+n = \frac1n$, for every positive integer $n$.
Draw an $x$-axis, and plot on it the first
ten points of the sequence.
What is $\lim_{n\goesto\infty} s_n$?

\ex{9.1.5}
Let $r$ be a real number,
and consider the sequence $1, r, r^2, r^3, \ldots.$
Show that the sequence converges if and only
if $-1 < r \leq 1$, and that
\[
\lim_{n\goesto\infty} r^n =
\trilemma{0 & \mbox{if $-1<r<1$,}}
{1 & \mbox{if $r=1$,}}
{\infty & \mbox{if $r>1$.}}
\]
What is the behavior of the sequence for
$r=-1$ and for $r<-1$?
(\emph{Hint:} Let $r^n = e^{n \ln r}$,
for $r>0$.)

\ex{9.1.6}
Finish the proof of Theorem \thref{9.1.1}:
\begin{exenum}
\x
Prove \thref{9.1.1ii}.
\x
Prove \thref{9.1.1iii}.
\x
Prove \thref{9.1.1iv}.
\end{exenum}

\ex{9.1.7}
Let $s$ and $t$ be two infinite sequences
and $a$ a real number such that
\[
s_n = a + t_n
,
\]
for every integer $n$ greater than or equal
to some integer $k$.  Prove that
\[
\lim_{n\goesto\infty} s_n =
a + \lim_{n\goesto\infty} t_n
.
\]
[\emph{Suggestion:} It is easy to prove this
result directly from the definition of convergence.
Alternatively, one may consider a constant sequence
with the single value $a$, and obtain the result
as a corollary of Theorem \thref{9.1.1i}.]

\ex{9.1.8}
Consider the sequence $\{ s_n \}$
defined by $s_n = n + (-1)^n$,
for every integer $n \geq 0$.
\begin{exenum}
\x
Write the first ten terms of the sequence.
\x
Show that $\lim_{n\goesto\infty} s_n = \infty$,
but that $\{ s_n \}$ is not an increasing sequence.
\x
Give another example of a sequence $\{ s_n \}$
which is not monotonic but for which
$\lim_{n\goesto\infty} s_n = \infty$.
\end{exenum} 

\end{exercises}
