\begin{exercises}

\ex{9.2.1}
Determine whether or not each of the following
infinite series converges, and evaluate it if it does.
\begin{exenum}
\x
$\sum_{i=0}^\infty \frac7{5^i}$
\x
$\sum_{k=1}^\infty \frac{a}{5^k}$
\x
$\sum_{n=1}^\infty \left( \frac1{2^n} + \frac1n \right)$
\x
$\sum_{j=0}^\infty \left(\frac1{2^j} - \frac1{3^j}\right)$
\x
$\sum_{i=1}^\infty \frac{5\cdot2^i+6i}{i2^i}$
\x
$\sum_{k=0}^\infty \left(3+\frac1{3^k}\right)$
\x
$\sum_{i=1}^\infty \frac{i^2-1}{i^2+1}$
\x
$\sum_{k=0}^\infty \frac{2^k+3^k}{6^k}$.
\end{exenum}

\ex{9.2.2}
Consider the infinite series
$\sum_{i=0}^\infty a_i$ defined by
\[
\cond{a_{2i} = \frac1{2^i}, & i = 0, 1, 2, \ldots ,}
\]
\[
\cond{a_{2i+1} = 0, & i = 0, 1, 2, \ldots .}
\]
Write out the sum of the first ten terms.
Does the series converge?
If so, to what value?

\ex{9.2.3}
Using Theorem \thref{9.2.2}, show that if
$\sum_{i=m}^\infty a_i$ converges and if
$\sum_{i=m}^\infty b_i$ diverges, then
$\sum_{i=m}^\infty (a_i+b_i)$ must diverge.

\ex{9.2.4}
Is it true that if the series
$\sum_{i=m}^\infty (a_i+b_i)$ converges, then both
$\sum_{i=m}^\infty a_i$ and
$\sum_{i=m}^\infty b_i$ must also converge?
Give a reason for your answer.

\ex{9.2.5}
Prove that the harmonic series
$\sum_{k=1}^\infty \frac1k$ diverges using
the following elementary argument.
Begin by grouping the terms of the series:
\[
\sum_{k=1}^\infty \frac1k =
1 + \left(\frac12+\frac13\right) + 
\left(\frac14+\frac15+\frac16+\frac17\right)
\]
\[
+ \left(\frac18+\frac19+\frac1{10}+\frac1{11}+\frac1{12}+
\frac1{12}+\frac1{13}+\frac1{14}+\frac1{15}\right) + \cdots
,
\]
and observe that
\[
\frac12 + \frac13 > \frac14 + \frac14 = \frac12
,
\]
\[
\frac14+\frac15+\frac16+\frac17>
\frac18+\frac18+\frac18+\frac18=\frac12,
\quad \mbox{etc.}
\]

\ex{9.2.6}
Consider the infinite series
$\sum_{k=1}^\infty \left(\frac1k - \frac1{k+1}\right)$.
By writing out a few terms of the sequence
of partial sums, show that the series converges,
and give its value.

\ex{9.2.7}
An infinite series of the form
$\sum_{i=m}^\infty (a_i - a_{i+1})$
is called a \dt{telescoping series}
(see Problem \exref{9.2.6}).
Prove that it converges if and only if the sequence
$\{ s_n \}$ converges.
If it does converge, what is its value?

\ex{9.2.8}
Determine whether or not each of the following infinite
series converges, and evaluate it if it does.
\begin{exenum}
\x
$\sum_{k=1}^\infty \frac1{k(k+1)}$.
\x
$\sum_{i=1}^\infty \frac{2i+1}{i^2(i^2+2i+1)}$.
\x
$\sum_{k=1}^\infty \ln \left(\frac{k+1}k\right) =
\ln \left(\frac21\right) + \ln \left(\frac32\right) +
\ln \left(\frac43\right) + \cdots $.
\end{exenum}
(\emph{Hint:} Look at Problems
\exref{9.2.6} and \exref{9.2.7}.)

\end{exercises}
